Statically Typed String Sanitation Inside a Python

### 2014-10-07 sanitation-psp14/sanitation-psp14-short.tex
Rearranged def 1
Final edits for PSP14

\begin{defn}[Definition of \lhead{r}]
The function $\lhead{r}$ is defined in terms of the structure of $r$:
\begin{itemize}[noitemsep]
\item $\lhead{a r'} = a$ where $a \in \Sigma$
\item $\lhead{r_1r_2} = \lhead{r_1}$
\item $\lhead{\epsilon} = \epsilon$
\item $\lhead{r_1 +  r_2}  = \lhead{r_1} + \lhead{r_2}$
\item $\lhead{r*} = \lhead{r}$
\item $\lhead{.} = a_1 + a_2 + ... + a_n$ for all $a_i \in \Sigma$ where $|\Sigma| = n$.
\end{itemize}
\end{defn}

@@

\begin{defn}[Definition of \ltail{r}]
The function \ltail{r} is defined in terms of $\lhead{r}$.
Note that $\lhead{r} = a_1 + a_2 + ... + a_i$.
We define $\ltail{r} = \delta_{a_1}(r) + \delta_{a_2}(r) + ... + \delta_{a_i}(r)$.
\end{defn}

@@

The rule 
 \textsc{S-T-SafeCoerce} checks for language inclusion, which we write $\lang{r_1} \subseteq \lang{r_2}$. Language inclusion is a decidable relation. As a result, the rule \textsc{S-E-SafeCoerce} does not need to perform any checks. The rule 
\textsc{S-T-Check} does not perform any static checks on the two languages involved, only checking that the two branches have the same type. The checks are performed dynamically, shown in the rules \textsc{S-E-Check-Ok} and \textsc{S-E-Check-NotOk}.

@@

We avoid the most significant issues inherent to proofs related to big-step semantics by avoiding non-termination in our specification.
The simply typed lambda calculus terminates, and our conservative extension clearly
terminates because new binding structures are not introduced and termination (i.e. decidability) of ${\sf subst}$ and ${\sf lreplace}$ is known. Even in a non-terminating calculus, type safety holds.
However, a more careful treatment or treatment for a language with non-termination would need to proceed by either a coinductive argument \cite{coinductionoo} or  with a small-step semantics.

@@

\begin{thm}[Safety for $\lambdap$] If $\emptyset \vdash \iota : \tau$ 
  then $\iota \sreduces \dot v$ and $\tctx \vdash \dot v : \tau$.
\end{thm}

@@

The translation correctness theorem guarantees that the translation is type preserving and that semantics of the original code and translation coincide, following the treatment of compilation pioneered by the TIL compiler for SML \cite{tarditi+:til-OLD}. We apply the translation inline in judgements for concision when convenient.

@@

\begin{thm}[Correctness of Input Sanitation for Translated Terms]\label{thm:main}
  If $\trden{e} = \iota$ and $\emptyset \vdash e : \stringin{r}$ then $\iota \sreduces \tstr{s}$
  for $s \in \lang{r}$.
\end{thm}
\begin{proof}
  By Theorem \ref{thm:trcorrect}, $\iota \sreduces \tstr{s}$ implies that $e \sreduces \strin{s}$.
  By Theorem \ref{thm:scorrect}, the above property together with the assumption that $e$ is well-typed implies that $s \in \lang{r}$.
\end{proof}

>>>>

\begin{defn}[Definition of \lhead{r}]
The relation $\lhead{r}=r'$ is defined in terms of the structure of $r$:
\begin{itemize}[noitemsep]
\item $\lhead{\epsilon} = \epsilon$
\item $\lhead{.} = a_1 + a_2 + ... + a_n$ for all $a_i \in \Sigma$ where $|\Sigma| = n$.
\item $\lhead{a} = a$ %where $a \in \Sigma$
\item $\lhead{r_1\cdot r_2} = \lhead{r_1}$
\item $\lhead{r_1 +  r_2}  = \lhead{r_1} + \lhead{r_2}$
\item $\lhead{r*} = \lhead{r}$
\end{itemize}
\end{defn}

@@

\begin{defn}[Definition of \ltail{r}]
The relation $\ltail{r} = r'$ is defined in terms of $\lhead{r}$.
Note that $\lhead{r} = a_1 + a_2 + ... + a_i$.
We define $\ltail{r} = \delta_{a_1}(r) + \delta_{a_2}(r) + ... + \delta_{a_i}(r)$.
\end{defn}

@@

The rule 
 \textsc{S-T-SafeCoerce} checks for language inclusion, which we write $\lang{r_1} \subseteq \lang{r_2}$. Language inclusion is a decidable relation. As a result, the rule \textsc{S-E-SafeCoerce} does not need to perform any checks. The rule 
\textsc{S-T-Check}, conversely, does not perform any static checks on the two languages involved, only checking that the two branches have the same type. The checks are performed dynamically, by rules \textsc{S-E-Check-Ok} and \textsc{S-E-Check-NotOk}.

@@

We avoid the most significant issues inherent to proofs related to big-step semantics by avoiding non-termination in our specification.
The simply typed lambda calculus terminates, and our conservative extension clearly
terminates because new binding structures are not introduced and termination (i.e. decidability) of ${\sf subst}$ and ${\sf lreplace}$ is known. Even in a non-terminating calculus, type safety would hold, but a more careful treatment or treatment for a language with non-termination would need to proceed by either a coinductive argument \cite{coinductionoo} or  with a small-step semantics.

@@

\begin{thm}[Safety for $\lambdap$] If $\emptyset \vdash \iota : \tau$ 
  then $\iota \sreduces \dot v$ and $\emptyset \vdash \dot v : \tau$.
\end{thm}

@@

The translation correctness theorem guarantees that the translation is type preserving and that semantics of the original code and translation coincide, following the treatment of compilation pioneered by the TIL compiler for SML \cite{tarditi+:til-OLD}. Note that we apply the translation inline in judgements for concision when convenient.

@@

\begin{thm}[Correctness of Input Sanitation for Translated Terms]\label{thm:main}
  If $\trden{e} = \iota$ and $\emptyset \vdash e : \stringin{r}$ then $\iota \sreduces \tstr{s}$
  for $s \in \lang{r}$.
\end{thm}
\begin{proof}
  By Theorem \ref{thm:trcorrect} and the rules given, $\iota \sreduces \tstr{s}$ implies that $e \sreduces \strin{s}$.
  Theorem \ref{thm:scorrect} together with the assumption that $e$ is well-typed implies that $s \in \lang{r}$.
\end{proof}
